\section*{Introduction}
\label{sec:introduction}
\addcontentsline{toc}{section}{Introduction}

Spoofax is a language workbench developed by Delft University of
Technology over the course of several years. During those years it has
grown to be \textit{``a language workbench for efficient, agile
  development of textual domain-specific languages with
  state-of-the-art IDE support''}~\cite{Kats10a}.

When developing new domain-specific languages using Spoofax, the
ability for rapid prototyping using a read-eval-print loop (REPL)
would be very convenient. This report describes the initial research
phase for this project: extending Spoofax with generating REPLs for the
languages defined with it.

The first three sections of this research report give an overview of
the problem domain. \Cref{sec:spoofax} gives the scope of Spoofax and
each of its meta-languages. In \cref{sec:repl}, REPLs are explained in
more detail. After that \cref{sec:literate-programming} introduces
literate programming.

The remaining four sections concern the problem and the project. In
\cref{sec:problem-definition}, a short problem definition is
given. \Cref{sec:problem-analysis} then gives an analysis of this
problem definition in the context of the services that Spoofax already
provides, to clarify potential problems and their solutions to
generate REPLs that work with languages defined in Spoofax. Next, in
\cref{sec:requirement-analysis}, a list of design goals will be
formalized to guide the development of the deliverable. From these
design goals and the problem definition and analysis, a list of
requirements for the product will be compiled. Lastly,
\cref{sec:realisation-product} gives an account of the development
frameworks and tools to be used during the development of the project.

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:
