\chapter{Discussion and Recommendations}
\label{cha:disc-recomm}

In this chapter the achieved results are discussed, as well as some tasks that
have not been achieved. \Cref{cha:evaluation} already discussed how the final
product satisfied the client's requirements. This chapter goes more in-depth why
not all these requirements have been met.

Afterwards, some recommendations are made that will further enhance and provide
possible directions for future work on the product.

\section{Discussion}
\label{sec:discuss-discussion}

TODO: section introduction

\input{discuss-recommend/dynsem}

%\input{discuss-recommend/eclipse}

\section{Recommendations}
\label{sec:discuss-future}

This section makes several recommendations to the client for future work on the
delivered product. These recommendations come forth from requirements that have
not been implemented (see \cref{ssec:eval-requirements}). First, recommendations
are made to provide additional frontend implementations. Afterwards, alternative
evaluation strategies are discussed to have the backend work with interpreters
other than DynSem.

\subsection{Alternative Frontend Implementations}
\label{ssec:discuss-alternative-frontend}

\Cref{sec:frontends} showed that implementing different frontends is
straightforward: only a handful of interfaces have to be implemented in order to
provide a basic frontend. The client has expressed their interest in numerous
other frontends, which have not been implemented due to time contraints.
However, because providing additional implementations is straightforward,
several recommendations for additional frontends are made in this subsection.

\input{discuss-recommend/intellij}

\input{discuss-recommend/literate-prog}

\subsection{Alternative Evaluation Strategies}
\label{ssec:discuss-alternate-eval}

\input{discuss-recommend/alternate-eval}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:
