\chapter{Evaluation}
\label{cha:evaluation}

Two evaluations are made in this chapter. The first section of this chapter,
\cref{sec:product-evaluation}, evaluates the product. This is done firstly by
looking at the requirements and secondly by looking at the design goals, both as
described in \cref{sec:requirement-analysis}. The end of this section concludes
with an evaluation of the product as a whole in \cref{ssec:eval-product}. The
second section of this chapter, \cref{sec:process-evaluation}, evaluates the
process by which the product came to be. It discusses the development
methodologies used and the feedback as received by the Software Improvement
Group.

\section{Product Evaluation}
\label{sec:product-evaluation}

\Cref{sec:requirement-analysis} describes the requirements and the design goals
that were agreed upon with the client at the start of the project. This section
evaluates how these are met by the product. \Cref{ssec:eval-requirements}
discusses each requirement and decides whether or not it has been met and how.
\Cref{ssec:eval-design-goals} does the same for the design goals.  Finally, this
section concludes with a short discussion to see if the minimal viable product
as defined in \cref{ssec:mvp} has been delivered.

\subsection{Evaluation of the requirements}
\label{ssec:eval-requirements}

This section discusses all the requirements as set out in
\cref{ssec:requirements}. All the ``must have'', ``should have'' and ``could
have'' requirements are listed in \cref{table:requirements}. Per requirement,
the table lists whether it has been met, including an explanation as to how it
has been met or why it has not been met.

\begin{table}[]
\centering
\resizebox{\textwidth}{!}{%
\begin{tabular}{|l||l||l|}
\hline
\textbf{Requirement} & \textbf{Met?} & \textbf{Explanation} \\ \hline
Interactive REPL     & Yes           & \begin{tabular}[c]{@{}l@{}}Two interactive REPLs have been created: one for the console\\
								  and another in Eclipse.\end{tabular}\\ \hline
\begin{tabular}[c]{@{}l@{}}Works with any language\\ defined in
Spoofax\end{tabular} & Yes           & \begin{tabular}[c]{@{}l@{}}The backend retrieves the information it needs from Spoofax's\\
								  services. A language developer is able to define REPL-specific\\
					  			  configuration in an ESV file, as well.\end{tabular} \\ \hline
Input history        & Yes           & \begin{tabular}[c]{@{}l@{}}An interface with a corresponding client-agnostic implementation\\
								  has been created in the backend, allowing frontends to either use\\
								  the implementation or implement the interface as they see fit.\\
				  				  Both the console and the Eclipse frontend provide input history.\end{tabular} \\ \hline
\begin{tabular}[c]{@{}l@{}}Automatic binding of\\
previously yielded values\end{tabular} & Yes & \begin{tabular}[c]{@{}l@{}}The language designer can specify these semantics in DynSem.\end{tabular} \\ \hline
Multiline input editing                & Yes & \begin{tabular}[c]{@{}l@{}}The way in which input is given is not enforced by the backend.\\
									  However, both the console and the Eclipse frontend implement\\
					  				  multiline input editing.\end{tabular}\\ \hline
Error reporting      & Yes           & \begin{tabular}[c]{@{}l@{}}Both the console and the Eclipse frontend print errors to the user.\\
								  The characters to which these errors belong are highlighted in red.\\
							  	  Error reporting is not yet enabled during typing.\end{tabular}\\ \hline
Syntax-checked expressions             & Yes & \begin{tabular}[c]{@{}l@{}}The backend parses the input with the services provided by Spoofax.\\
							  		  As such, syntax-checked input was given for free. However, this\\
								  	  feedback is not yet enabled during typing.\end{tabular}\\ \hline
Syntax highlighting & No             & \begin{tabular}[c]{@{}l@{}}A working prototype has been implemented. Due to the nature of\\
								  console-based user interfaces, the console frontend cannot support\\
								  this feature. The Eclipse frontend has most of the required\\
							   	  functionality.\end{tabular}\\ \hline
Integration with Eclipse               & Yes & \begin{tabular}[c]{@{}l@{}}A REPL has been implemented in Eclipse, allowing the user to give\\
							  		  input and see output.\end{tabular}\\ \hhline{|=|=|=|}
Ability to redefine identifiers        & Yes & \begin{tabular}[c]{@{}l@{}}The language designer can specify these semantics in DynSem.\end{tabular} \\ \hline
Environment inspection                 & No  & \begin{tabular}[c]{@{}l@{}}This is impossible to do in the current DynSem implementation,\\
								  	  because the environment is captured in an object of type \texttt{Object}\\
								  	  and is therefore unprintable.\end{tabular} \\ \hline
Save and load REPL state               & No  & \begin{tabular}[c]{@{}l@{}}Due to limitations in time, this feature has not been implemented.\end{tabular}\\ \hhline{|=|=|=|}
Syntactic code completion              & No  & \begin{tabular}[c]{@{}l@{}}Spoofax's code completion services do not yet work with all\\
								  	  languages and are in process of a rewrite. As such, code\\
								 	  completion has not been attempted.\end{tabular}\\ \hline
\begin{tabular}[c]{@{}l@{}}Hover over variables to see\\
value, type and others\end{tabular}    & No  & \begin{tabular}[c]{@{}l@{}}Due to limitations in time and the fact that the environment is\\
					    			          captured in an object of type \texttt{Object}, this feature has\\
									  not been implemented.\end{tabular}\\ \hline
Literate programming                   & No  & \begin{tabular}[c]{@{}l@{}}An attempt has been made at writing an IPython frontend for the\\
								 	  REPL backend. However, due to limitations in time and the pressure\\
								  	  of finishing more important features, this work has not been finished.\end{tabular} \\ \hline
\begin{tabular}[c]{@{}l@{}}Integration with other\\
IDEs (IntelliJ)\end{tabular}           & No  & \begin{tabular}[c]{@{}l@{}}The Eclipse frontend is not entirely done yet, which was a\\
									  requirement before starting on implementing a frontend for\\
									  IntelliJ. As such, this feature has not been attempted.\end{tabular}\\ \hline
\end{tabular}
}
\caption{Evaluation of the requirements, separated per MoSCoW category.}
\label{table:requirements}
\end{table}

\subsection{Evaluation of the design goals}
\label{ssec:eval-design-goals}

In this section the design goals as set out in \cref{ssec:goals} are
discussed. Each design goal is listed below, along with an explanation as to how
it has been met.

\paragraph{Language-agnostic} This design goal stated that the REPL should work
for any language. Additionally, the language designer should be provided with a
language-agnostic interface to configure and implement the language-specific
parts of the REPL.

An extension of Spoofax's Editor Services has been created, allowing language
designers to configure certain parts of the REPL. Additionally,
language-specific changes, such as the ability to redefine identifiers, can be
made through DynSem. Currently, however, only DynSem-based interpreters are
supported, as opposed to Java- and Stratego-based interpreters. Despite this,
care has been taken to make this easily extendable. This design goal, therefore,
is met.

\paragraph{Minimal configuration} This design goal stated that the REPL should
require as little configuration as possible by reusing existing components of
the language definitions and automatically deriving other settings.

The extension of the Editor Services discussed in the previous design goal is
minimal. Three settings can be configured at the time of writing, all of which
have sane default settings. One additional setting is mandatory and cannot be
derived automatically (see \cref{ssec:supp-diff-ways}): the name of the
evaluation strategy to use. Additionally, because the existing DynSem
specification can be reused, only a minimal amount of new DynSem rules have to
be written to specify REPL-specific semantics. Finally, the SDF3 configuration
is reused to specify which start symbols are valid in case the language designer
wants to allow different start symbols. Therefore this design goal is met.

\paragraph{Maintainability} This design goal stated that the code should be
maintainable, because ownership will be transferred to the people already
working on Spoofax.

Various static analysis tools, up-to-date documentation, proper software design
and a received 4,5 out of 5 stars during the first round of feedback from the
Software Improvement Group (see \cref{ssec:sig} below), are good indications that
this design goal is met.

\paragraph{IDE-agnostic} This design goal stated that the REPL should place no
assumptions on the environment it runs in, since an effort is underway to do the
same for Spoofax itself.

The backend is agnostic to any frontend, no assumptions are made whatsoever:
console or graphical, blocking or unblocking user input, single- or
multithreaded, et cetera. This design goals is met.

\paragraph{Performance} This design goal stated that the REPL should be
performant, because care is taken to ensure that Spoofax itself is performant.

Launching the REPL is near instantaneous. Commands and expressions are evaluated
in real time, without noticeable delay. This design goals is thus met.

\paragraph{Modify Spoofax's existing codebase as little as possible} This design
goal stated that the REPL should be an extension to Spoofax, which means that
the existing codebase should be modified as little as possible.

The REPL is developed into a standalone repository. It uses the same API exposed
by Spoofax as any other extension would use. Minimal changes have been made to
Spoofax itself: the only large change that went upstream is the ability to
configure the REPL in an ESV file. However, DynSem had to be significantly
modified after its rewrite to support the use case of a REPL. These changes were
made in cooperation with its maintainer, and most were features that had to be
implemented regardless. As such, this design goal is also met.

\subsection{Evaluation of the product}
\label{ssec:eval-product}

This section evaluates whether the minimal viable product, as defined in
\cref{ssec:mvp}, has been delivered. The minimal viable product has been defined
as all the ``must have'' requirements (see \cref{ssec:requirements}), including
the design goals of being language-agnostic and having a minimal configuration.

The previous section concluded that all the design goals have been met.
\Cref{ssec:eval-requirements} and \cref{table:requirements} show that all the
``must have'' requirements have been implemented, except one: syntax
highlighting. The backend is ready to provide this functionality, and a working
prototype has been developed.

It is currently impossible to provide syntax highlighting in the console
frontend. The text-only environment is the limiting factor: once text has been
printed, it cannot be changed. This means that text would have to be printed
twice: once as it is typed in by the user, and again with syntax highlighting
applied.

The Eclipse frontend has the required functionality to provide syntax
highlighting. The backend, however, has no way of telling which widget (see
\cref{fig:frontend-eclipse}) to send the resulting text to. This means that as
of yet, only text in the output buffer can be highlighted.

As one might conclude, much of the code to implement syntax highlighting is
present. Finishing this work would not take much more effort. As such, this must
have requirement is close to being met. The minimal viable product is therefore
deemed to be delivered.

\section{Process Evaluation}
\label{sec:process-evaluation}

This section gives an evaluation of the process by which the delivered product
came to be. First, the methodologies used during development are discussed in
\cref{ssec:dev-meth}. \Cref{ssec:sig} shows the feedback received from the
Software Improvement Group and discusses the changes that were made to the code
accordingly.

\subsection{Development methodologies}
\label{ssec:dev-meth}

In the beginning of the project, agreements were made on the development
methodologies that were to be used. Throughout the duration of the project, some
of these methodologies turned out to work well, whilst others were less
effective than anticipated. This section discusses these agreements.

The Scrum methodology was chosen to manage the product development. Previous
experience showed that one week sprints are often too short, so for this project
biweekly sprints were decided. This worked well: there was enough time to make
informed decisions and to evaluate or rework certain things. Biweekly sprints
also allow for an easier revision for the sprint plan if changes need to be
made due to unexpected issues.

Sprints were evaluated with a meeting with the client, during which the progress
was demonstrated. The feedback received during these meetings was valuable and taken
into account to guide the next sprint.

The client mandated the use of Trello to manage the sprints, which has been a
positive experience as well. Trello makes it easy to manage deadlines and to
know what has to be done at what time. Managing a backlog of user stories and
moving them around from, for example, ``developing'' to ``testing'', is
straightforward and helps to keep everyone, including the client, updated on the
progress.
The pull-based development model worked well, as was anticipated from previous
experience. The continuous integration, in combination with a set of strict
static analysis tools, ensured that the code review process had a positive
influence on the quality of the code.\\

Naturally, there were also things that could have gone better. It is easy to
underestimate the amount of work when putting together a sprint plan, which has
been a source of problems. The solution for this issue is to deliberately plan
less: it is easier to move new user stories to the sprint plan than it is to
decide which user stories not to complete.

A second issue concerns the unstable development environment provided by Eclipse.
Omitting all the details, several hard to track issues occurred which took a
considerable amount of time to resolve. To add to this, on several occurrences
the development of the Eclipse plugin had to wait for changes to make it into
Spoofax.

The biggest issue that occurred during the development is the fact that Spoofax
is a moving target. Halfway through the allotted time for development, DynSem
was significantly rewritten. After this rewrite, the functionality required for
a REPL was no longer present. To resolve this, considerable effort has been spent
to add these features to DynSem in cooperation with DynSem's maintainer. All of
this work has been accepted upstream and has now been published in the latest
snapshot releases.\\

To conclude, the process by which the product came to be went predominantly
well, despite some issues that can be improved during future projects.

\subsection{Software Improvement Group}
\label{ssec:sig}

The code was sent to the Software Improvement
Group\footnote{\url{https://www.sig.eu/}} for review on May 27th. The following
feedback was received on the third of June:

\begin{quotation}
[Analyse]

De code van het systeem scoort 4,5 ster op ons onderhoudbaarheidsmodel, wat
betekent dat de code bovengemiddeld onderhoudbaar is. De hoogste score is niet
behaald door een lagere score voor Unit Interfacing.

Voor Unit Interfacing wordt er gekeken naar het percentage code in units met een
bovengemiddeld aantal parameters. Doorgaans duidt een bovengemiddeld aantal
parameters op een gebrek aan abstractie. Daarnaast leidt een groot aantal
parameters nogal eens tot verwarring in het aanroepen van de methode en in de
meeste gevallen ook tot langere en complexere methoden.

In jullie geval heeft de constructor van \texttt{Data} een enorm aantal
parameters. Er zijn grofweg twee manieren om dat te addresseren: de eerste is
het groeperen van velden in sub-types. Een flauw voorbeeld is het groeperen van
\texttt{width} en \texttt{height} in een type \texttt{Dimensions}.
Dit voorbeeld is natuurlijk triviaal, maar je ziet wel vaak dat studenten aan
het begin wat huiverig zijn om kleine classes voor alleen een type te
introduceren, terwijl dit op de lange termijn vaak wel degelijk zin heeft. De
tweede manier is niet meer alle parameters in de constructor meegeven, maar via
setter-methodes. Dat leidt tot meer code, maar maakt het intialiseren van een
object wel leesbaarder.

Een ander voorbeeld van dezelfde situatie is de constructor van
\texttt{TransformCommand} (zo te zien krijgen jullie daar ook al een
waarschuwing vanuit CheckStyle). Dit geval is wat moeilijker op te lossen,
aangezien jullie hier van dependency injection gebruik maken en de methode
daardoor wat moeilijker kunnen refactoren.

Het is goed om te zien dat jullie naast productiecode ook veel testcode hebben
geschreven. De verhouding tussen de hoeveelheid productiecode en testcode is met
bijna 1 op 1 ook zeer gezond. Hopelijk lukt het jullie om dat vast te houden
tijdens het vervolg van het project.
\end{quotation}

This analysis says that the code scored bad on unit interfacing, because there
were quite a few classes that had an above average amount of constructor
parameters. The reason that this came to be is the fact that dependency
injection is used, as opposed to the traditional way of managing dependencies.
This fact is also mentioned in the analysis itself.

Two suggestions were given as a possible solution to this problem: create new
classes that encapsulate several parameters, or use \texttt{setter} methods
to pass parameters this way. However, neither solution was satisfactory as-is:
data classes are often considered to be a code smell, and \texttt{setter}
methods risk not fully initializing an object. The classes to which the
criticism applied have been refactored into smaller, more manageable units.
Ultimately, this not only led to resolving the only negative feedback made by
the Software Improvement Group, it also led to a better architecture.

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:
