\section{Dependency Injection}
\label{sec:injection}

The design of Spoofax makes extensive use of dependency injection, specifically
by using the Guice framework\footnote{\url{https://github.com/google/guice/wiki/Motivation}}.
To achieve as much interopability as possible, the final product
uses dependency injection in the same manner as Spoofax. Guice had a
substantial impact on the design and implementation of the final product, since
it nearly eliminates the need to create factories for complex datatypes.

Guice requires the programmer to bind implementations of dependencies in its
module classes, thereby separating behavior and dependency resolution.  Complex
datatypes can then accept their dependencies as constructor arguments, allowing
Guice to resolve and inject an instance of a bound type.

In the final product dependency injection is also used to supply classes
with a default configuration.  An example is the default list of commands
available to a user as described in \cref{sec:commands}.  The list of commands
is created from a Guice module by instantiating a \texttt{MapBinder} with
predefined bound commands. Child modules can then append extra entries to the
\texttt{MapBinder}, which makes them directly available to the existing
architecture.

Using dependency injection has made it easier to create a modular product,
centered around smaller interfaces interacting with each other. All these
interfaces can easily be bound to new implementations, thereby extending the
functionality of the REPL.
